---
title: "Confirmatory Data Analysis (CDA)"
author: "Jia Peng Chua | Zanelle Lee Wenfang"
date-modified: "last-modified"
execute:
    echo: true
    eval: true
    warning: false
    freeze: true
---

# 1. Getting Started

In this section, we conduct confirmatory data analysis of dengue cases with factors such as gender, age groups, imported vs local cases, dengue serotypes and Taiwan counties.

## 1.1 Loading R Packages

The following R packages will be required:

-   `tidyverse`: Used for data manipulation, visualisation and analysis

-   `plotly`: For plotting interactive statistical graphs

-   `sf`: For analysing and visualising spatial data

-   `tmap`: Create static and interactive thematic maps

-   `GGally`: Provide additional visualisation functions

-   `ggstatplot`: For integrating statistical analysis into visualisations

-   `ggmosaic`: For visualisation of categorical data relationships and associations

```{r}
#| code-fold: true
#| code-summary: "Loading R packages"
pacman::p_load(tidyverse, plotly, sf, tmap, GGally, ggstatsplot, ggmosaic)
```

## 1.2 Importing Dengue Dataset

```{r}
#| code-fold: true
#| code-summary: "Importing Dengue_Daily dataset"
dengue_daily <- read_csv("data/dengue_daily_en.csv")

summary(dengue_daily)
```

### 1.3 Additional Tables Required for the Analysis

```{r}
#| code-fold: true
#| code-summary: "Show code"

group_by_age <- dengue_daily %>%
  group_by(Age_Group, Onset_Year) %>%
  summarize(Count = n(), .groups = "drop")

group_by_gender <- dengue_daily %>%
  group_by(Gender, Onset_Year) %>%
  summarize(Count = n(), .groups = "drop")

group_by_serotype <- dengue_daily %>%
  group_by(Serotype, Onset_Year) %>%
  summarize(Count = n(), .groups = "drop")

group_by_county <- dengue_daily %>%
  group_by(Residential_County_City, Onset_Year) %>%
  summarize(Count = n(), .groups = "drop")

group_by_imported <- dengue_daily %>%
  group_by(Imported_Case, Onset_Year) %>%
  summarize(Count = n(), .groups = "drop")
```

# 2. Confirmatory Data Analysis (CDA)

## 2.1 Does certain age group have significantly higher dengue cases compared to others?

First, let's visualise the distribution of mean dengue cases by age groups.

```{r}
#| code-fold: true
#| code-summary: "Show code"

mean_age_group_year <- dengue_daily %>%
    group_by(Age_Group, Onset_Year) %>%
    summarize(Mean = mean(n()), .groups = "drop")

ggplot(mean_age_group_year, aes(x = Age_Group, y = Mean, fill = Age_Group)) +
    geom_boxplot() +
    labs(title = "Distribution of Dengue Cases Across Age Groups",
         x = "Age Group",
         y = "Mean Dengue Cases per Year")
```

### 2.1.1 Check if dengue cases are evenly distributed across age groups

We perform chi-square test using the sum of dengue cases by age group.

-   H~0~: The number of dengue cases is independent of age group

-   H~1~: The number of dengue cases depends on age group

::: panel-tabset
#### Chi-Square test

```{r}
#| code-fold: true
#| code-summary: "Show code"
sum_age_group <- dengue_daily %>%
    group_by(Age_Group) %>%
    summarize(Count = n(), .groups = "drop")

chisq.test(sum_age_group$Count)
```

#### Visual

```{r}
#| code-fold: true
#| code-summary: "Show code"
ggbetweenstats(
    data = sum_age_group,
    x = Age_Group, 
    y = Count,
    type = "np",
    mean.ci = TRUE, 
    p.adjust.method = "fdr",
    messages = FALSE
    )
```
:::

Conclusion:

-   Since p-value is \< 0.05, we reject the null hypothesis at 5% significance level.

-   This suggests that the number of dengue cases varies significantly across different age groups.

### 2.1.2 Check if dengue cases are evenly distributed across age groups over the years

We perform chi-square test using the sum of dengue cases by age group.

-   H~0~: The distribution of dengue cases across age group does not change over the years

-   H~1~: The distribution of dengue cases across age group changes over the years

::: panel-tabset
#### Chi-square test

```{r}
#| code-fold: true
#| code-summary: "Show code"
sum_age_group_year <- dengue_daily %>%
    group_by(Age_Group, Onset_Year) %>%
    summarize(Count = n(), .groups = "drop")

sum_age_group_year_table <- table(sum_age_group_year$Age_Group, 
                                  sum_age_group_year$Onset_Year)

chisq.test(sum_age_group_year_table)
```

#### Visual

```{r}
#| code-fold: true
#| code-summary: "Show code"
ggbetweenstats(
    data = sum_age_group_year,
    x = Age_Group, 
    y = Count,
    type = "p",
    mean.ci = TRUE, 
    pairwise.comparisons = TRUE, 
    pairwise.display = "s",
    p.adjust.method = "fdr",
    messages = FALSE
    )
```
:::

Conclusion:

-   Since p-value is \> 0.05, we do not reject the null hypothesis at 5% significance level.

-   This suggests that the number of dengue cases does not change across different age groups over the years.

## 2.2 How do other factors such as gender, imported cases, serotype and geographic distribution influence the number of dengue cases?

### 2.2.1 Gender

-   H~0~: There is no difference in the number of dengue cases between male and female

-   H~1~: There is a difference in the number of dengue cases between male and female

::: panel-tabset
#### T-test

```{r}
#| code-fold: true
#| code-summary: "Show code"
group_by_gender_test <- group_by_gender %>%
    filter(Gender %in% c("F", "M"))

t_test_gender <- t.test(Count ~ Gender, data = group_by_gender_test)

t_test_gender
```

#### Two-sample mean test

```{r}
#| code-fold: true
#| code-summary: "Show code"
ggbetweenstats(
    data = group_by_gender_test,
    x = Gender, 
    y = Count,
    type = "np",
    messages = FALSE
    )
```
:::

Conclusion:

-   Since p-value is \> 0.05, we do not reject the null hypothesis at 5% significance level.

-   This suggests that there is no difference in the number of dengue cases between male and female.

### 2.2.2 Imported Cases

-   H~0~: There is no difference in the number of dengue cases between imported and local cases

-   H~1~: There is a difference in the number of dengue cases between imported and local cases

::: panel-tabset
#### T-test

```{r}
#| code-fold: true
#| code-summary: "Show code"
t_test_gender <- t.test(Count ~ Imported_Case, data = group_by_imported)

t_test_gender
```

#### Two-sample mean test

```{r}
#| code-fold: true
#| code-summary: "Show code"
ggbetweenstats(
    data = group_by_imported,
    x = Imported_Case, 
    y = Count,
    type = "np",
    messages = FALSE
    )
```
:::

Conclusion:

-   Since p-value is \< 0.05, we reject the null hypothesis at 5% significance level.

-   This suggests that there is a difference in the number of dengue cases between imported and local cases.

### 2.2.3 Serotype

-   H~0~: There is no difference in the number of dengue cases between different serotypes

-   H~1~: There is a difference in the number of dengue cases between different serotypes

::: panel-tabset
#### Kruskal-Wallis test

```{r}
#| code-fold: true
#| code-summary: "Show code"

# Excluding data with missing Serotype
group_by_serotype_test <- group_by_serotype %>% filter(!is.na(Serotype))

kruskal.test(Count ~ Serotype, data = group_by_serotype_test)
```

#### Visual

```{r}
#| code-fold: true
#| code-summary: "Show code"
ggbetweenstats(
    data = group_by_serotype_test,
    x = Serotype, 
    y = Count,
    type = "p",
    mean.ci = TRUE, 
    pairwise.comparisons = TRUE, 
    pairwise.display = "s",
    p.adjust.method = "fdr",
    messages = FALSE
)
```
:::

Conclusion:

-   Since p-value is \< 0.05, we reject the null hypothesis at 5% significance level.

-   This suggests that there is a difference in the number of dengue cases between different serotypes.

### 2.2.4 Geographic Distribution

-   H~0~: There is no difference in the number of dengue cases between county cities

-   H~1~: There is a difference in the number of dengue cases between county cities

::: panel-tabset
#### Kruskal Wallis test

```{r}
#| code-fold: true
#| code-summary: "Show code"

kruskal.test(Count ~ Residential_County_City, data = group_by_county)
```

#### Visual

```{r}
#| code-fold: true
#| code-summary: "Show code"
ggbetweenstats(
    data = group_by_county,
    x = Residential_County_City, 
    y = Count,
    type = "p",
    mean.ci = TRUE, 
    pairwise.comparisons = TRUE, 
    pairwise.display = "s",
    p.adjust.method = "fdr",
    messages = FALSE
)
```
:::

Conclusion:

-   Since p-value is \< 0.05, we reject the null hypothesis at 5% significance level.

-   This suggests that there is significant difference in the number of dengue cases between county cities.

# 3. CDA Summary

+----------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| CDA            | Key Findings                                                                                                                                                                                                               |
+================+============================================================================================================================================================================================================================+
| Age Group      | -   Visually, the number of dengue cases varies significantly across different age groups. However, over the years, the number of dengue cases were generally stable except for the few years with spikes in dengue cases. |
|                |                                                                                                                                                                                                                            |
|                | -   From the statistical test, we conclude that the number of dengue cases varies significantly across different age groups.                                                                                               |
|                |                                                                                                                                                                                                                            |
|                | -   When we aggregate the data by year, the number of dengue cases does not change significantly across different age groups over the years.                                                                               |
+----------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Gender         | -   Visually, the distribution of dengue cases between male and female seems indifferent.                                                                                                                                  |
|                |                                                                                                                                                                                                                            |
|                | <!-- -->                                                                                                                                                                                                                   |
|                |                                                                                                                                                                                                                            |
|                | -   From the t-test, there is no difference in the number of dengue cases between male and female.                                                                                                                         |
+----------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Imported Cases | -   From the box plot, there are significantly more local cases than imported cases.                                                                                                                                       |
|                |                                                                                                                                                                                                                            |
|                | -   This is supported by t-test, where there is a difference in the number of dengue cases between imported and local cases.                                                                                               |
+----------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Serotype       | -   Visually, there is some difference between the means of the number of dengue cases across different serotypes.                                                                                                         |
|                |                                                                                                                                                                                                                            |
|                | -   This is supported by the Kruskal-Wallis test, where there is a difference in the number of dengue cases between different serotypes.                                                                                   |
+----------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| County Cities  | -   Visually, there is a significant difference in the number of dengue cases between county cities.                                                                                                                       |
|                |                                                                                                                                                                                                                            |
|                | -   This is supported by the Kruskal-Wallis test, where there is a difference in the number of dengue cases between different county cities.                                                                               |
+----------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
