---
title: "Data Preparation"
author: "Jia Peng Chua | Zanelle Lee Wenfang"
date-modified: "last-modified"
execute:
    echo: true
    eval: true
    warning: false
    freeze: true
---

# 1 Getting Started

The following R packages will be required:

```{r}
pacman::p_load(tidyverse,plotly)
```

# 2 Importing Data

The code chunk below uses `read_csv()` of `readr` to import the `dengue_daily` dataset into R environment.

```{r}
dengue_daily <- read_csv("data/Dengue_Daily.csv")
```

```{r}
summary(dengue_daily)
```

Translate column names to English

-   發病日: Onset Date

-   個案研判日: Case Classification Date

-   通報日: Reporting Date

-   性別: Gender

-   年齡層: Age Group

-   居住縣市: Residential County/City

-   居住鄉鎮: Residential Town/District

-   居住村里: Residential Village

-   最小統計區: Smallest Statistical Area

-   最小統計區中心點X: X-coord

-   最小統計區中心點Y: Y-coord

-   一級統計區: Primary Statistical Area

-   二級統計區: Secondary Statistical Area

-   感染縣市: Infection County/City

-   感染鄉鎮: Infection Town/District

-   感染村里: Infection Village

-   是否境外移入: Imported Case

-   感染國家: Infection Country

-   確定病例數: Confirmed Cases

-   居住村里代碼: Residential Village Code

-   感染村里代碼: Infection Village Code

-   血清型: Serotype

-   內政部居住縣市代碼: MOI Residential County Code

-   內政部居住鄉鎮代碼: MOI Residential Town Code

-   內政部感染縣市代碼: MOI Infection County Code

-   內政部感染鄉鎮代碼: MOI Infection Town Code

```{r}
colnames(dengue_daily) <- c("Onset_Date", "Case_Classification_Date", "Reporting_Date", "Gender", 
                  "Age_Group", "Residential_County_City", "Residential_Town_District", 
                  "Residential_Village", "Smallest_Statistical_Area", "X_coord", 
                  "Y_coord", "Primary_Statistical_Area", "Secondary_Statistical_Area",
                  "Infection_County_City", "Infection_Town_District", "Infection_Village",
                  "Imported_Case", "Infection_Country", "Confirmed_Cases", 
                  "Residential_Village_Code", "Infection_Village_Code", "Serotype", 
                  "MOI_Residential_County_Code", "MOI_Residential_Town_Code", 
                  "MOI_Infection_County_Code", "MOI_Infection_Town_Code")

```

## 2.1 Filter Past 10 Years' Data

By "發病日" column

```{r}
dengue_10yrs <- dengue_daily[dengue_daily$Onset_Date >= as.Date("2015-01-01") & dengue_daily$Onset_Date <= as.Date("2024-12-31"), ]
```

```{r}
summary(dengue_10yrs)
```

## 2.2 Check Missing Values

```{r}
sum(is.na(dengue_10yrs))
```

\<TBC\>

## 2.3 Data Cleaning

# 3. Data Preparation

Check Age_Group column

```{r}
age_group_count <- dengue_10yrs %>%
  count(Age_Group)

print(age_group_count)
```

Combine values 0,1,2,3,4, and sort the columns

```{r}
dengue_10yrs <- dengue_10yrs %>%
  mutate(Age_Group = ifelse(Age_Group %in% c("0", "1", "2", "3", "4"), "0-4", Age_Group),
         Age_Group = factor(Age_Group, levels = c("0-4", "5-9", 
                                                  sort(unique(Age_Group[
                                                      Age_Group != "0-4" & Age_Group != "5-9"]
                                                      )))))
```

```{r}
ggplot(data = dengue_10yrs, aes(x = Age_Group)) +
    geom_bar(color = "black", fill = "grey") +  
    ggtitle("Distribution of Age Groups") +
    xlab("Age Group") +
    ylab("Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

# 4. EDA

Aggregate by Onset_Month, count,

```{r}

dengue_10yrs <- dengue_10yrs %>%
  mutate(Date = ymd(Onset_Date))  # Convert Y/M/D to Date (use lubridate's ymd function)

# Extract the year from the Date column
dengue_10yrs <- dengue_10yrs %>%
  mutate(Onset_Year = year(Date))

aggregated_data <- dengue_10yrs %>%
  group_by(Age_Group, Residential_County_City, Onset_Year) %>%
  summarize(Count = n(), .groups = "drop")

cases_by_age <- aggregated_data %>%
  plot_ly(x = ~Age_Group, 
          y = ~Count, 
          color = ~Residential_County_City,
          text = ~Residential_County_City, 
          hoverinfo = "text",
          type = 'scatter',
          mode = 'markers',
          frame = ~Onset_Year) %>%
  layout(
    title = "Dengue Cases by Age Group",
    xaxis = list(title = "Age Group"),
    yaxis = list(title = "Count of Cases"),
    showlegend = FALSE
  )

cases_by_age
```
